---
title: "Rmd Exercise Template"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the data
load the gapminder dataset 

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```

## If Statement
report if there are any records from 2002 and 2012
```{r}
year<-2002
if(any(gapminder$year == year)){
   print(paste("Year", year,"found in the dataset."))
} else {
  print(paste("Year", year,"not found in the dataset."))
}

year<-2012
if(any(gapminder$year == year)){
   print(paste("Year", year,"found in the dataset."))
} else {
  print(paste("Year", year,"not found in the dataset."))
}

```

## Loop and If statement 
Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

```{r}
mean_pop <- mean(gapminder$pop)

for (i in unique(gapminder$country)) {
   mean_country <- mean(gapminder$pop[gapminder$country==i])
   
   if (mean_country < mean_pop) {
     mean_le <- mean(gapminder$lifeExp[gapminder$country==i])
     print(paste("The mean life expectancy of", i, "is", mean_le))
   } 
} 

```


Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r}
for (i in unique(gapminder$continent)){
    mean_le <- mean(gapminder$lifeExp[gapminder$continent == i])
    
    if (mean_le < 50)
        print(paste("Average life expectancy in", i, "is less than 50" ))
    else if (mean_le >= 50 & mean_le <= 70)
        print(paste("Average life expectancy in", i, "is between 50 and 70"))
    else
        print(paste("Average life expectancy in", i, "is greater than 70"))
}
 

```

## Writing Functions
Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset. Hint: Use mode() or class() to get the class of the data in each column. Remember that names() or colnames() returns the name of the columns in a dataset.  

```{r}
print_df_info <- function(df) {
  col_names <- colnames(df)
  for (i in col_names) {
    print(paste0(i, ": ", class(df[, i])))
  }
}
print_df_info(gapminder)
```


Create a function that given a vector will print the mean and the standard deviation of a vector, it will optionally also print the median. Hint: include an argument that takes a boolean (TRUE/FALSE) operator and then include an if statement.  
```{r}
print_vec_info <- function(vecname, print_med = FALSE){
    mean_vec <- mean(vecname)
    sd_vec <- sd(vecname)
    med_vec <- median(vecname)
    
    print(paste("The mean is", mean_vec))
    print(paste("The standard deviation is", sd_vec))
    
    if (print_med)
        print(paste("The median is", med_vec))
}

print_vec_info(gapminder$lifeExp, TRUE)
```

##Analyzing the relationship between GDP per capita and life expectancy
##Use what you???ve learned so far to answer the following questions using the ##gapminder dataset. Be sure to include some visualizations!

What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

```{r}
years <- unique(gapminder$year)
viz_lm <-
  function(df, y, x, year) {
    df_year <- df[df[["year"]] == year, ]
    logy <- log(df_year[[y]])
    logx <- log(df_year[[x]])
    fit <- lm(logy ~ logx)
    plot(logy ~ logx, main = year,
         xlab = x, ylab = y)
    lines (logx, predict(fit), col = "red")
  }
for (i in years) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)
}


```


Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.

```{r}
gapminder$new_continent <- gapminder$continent
gapminder$new_continent[gapminder$continent == "Americas"] <- "South America"
northamerica <- c("United States", "Mexico", "Canada")
gapminder$new_continent[gapminder$country %in% northamerica] <- "North America"
continent <- unique(gapminder$new_continent)

viz_lm <-
  function(df, y, x, continent) {
    df_cont <- df[df[["new_continent"]] == continent, ]
    logy <- log(df_cont[[y]])
    logx <- log(df_cont[[x]])
    fit <- lm(logy ~ logx)
    plot(logy ~ logx, main = continent,
         xlab = x, ylab = y)
    lines (logx, predict(fit), col = "red")
  }


for (i in continent) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)
}
```

