---
title: "R Day 2 exercises"
author: "Duyun Tan "
date: "`r Sys.Date()`"
output: html_document
---

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

<!-- BEFORE YOU START, REMEMBER TO RENAME THE .RMD FILE AND SAVE IT AS day2_exercises_LastNameFirstName.Rmd AND THEN REMOVE THIS INSTRUCTION -->

## Load the data
```{r}
library(dplyr)
library(data.table)
library(lubridate)
```

Read both California energy datasets. Make sure the `datetime` variable is in an appropriate data type (i.e. not character). 

```{r}
generation_dp <- read.csv(here::here("data/ca_energy_generation.csv"), 
                          stringsAsFactors = F)
imports_dp <- read.csv(here::here("data/ca_energy_imports.csv"), 
                       stringsAsFactors = F)
```

```{r}
generation_dp %>% mutate(datetime = as_datetime(datetime))
imports_dp %>%  mutate(datetime = as_datetime(datetime))

head(generation_dp)
head(imports_dp)
```

## Merge and reshape the data

Merge the two datasets and then melt the resulting dataframe/datatable to make it tidy.

```{r}
merged_two <- merge(generation_dp, imports_dp, by = "datetime")
head(merged_two)

long_ca_energy <- melt(merged_two, 
                       id.vars = "datetime",
                       variable.name = "source",
                       value.name = "output")

head(long_ca_energy)
```


## Creating new variables

Create a series of new variables: 

1. `day`, which is the year-month-day, without the hour. The `lubridate` function `as_date` will do this.
2. `log_output`, which is the natural log of the output.
3. **Challenge**: `per_output`, which is the percent of daily output represented by each observation. You will need to use `group_by` and to create a new variable with the total output for the day. (Make sure to use `ungroup()` after this!)

Bonus: If you are using `dplyr`, try to do this all in one pipe!

```{r}
names(long_ca_energy)

long_ca_energy <- long_ca_energy %>% 
    mutate(day = as_date(datetime), log_output = log(output)) %>% 
    group_by(day) %>% 
    mutate(total_output = sum(output)) %>% 
    ungroup() %>% 
    mutate(per_output = output/total_output)
    

long_ca_energy %>% select(day, log_output, per_output) %>% head()
```


## Summarizing and analyzing data

1. Which source has the greatest mean output by hour? (Hint: Use the `dplyr` verb `arrange(desc(variable))` to order the data frame so that the largest value of `variable` is first. Don't use `desc` and it arranges in ascending order. The `data.table` function is `setorder`.) Which has the least?
```{r}
long_ca_energy %>%
  group_by(source) %>%
  summarize(mean_hourly = mean(output, na.rm = T)) %>%
  arrange(desc(mean_hourly))

long_ca_energy %>%
  group_by(source) %>%
  summarize(mean_hourly = mean(output, na.rm = T)) %>%
  arrange(mean_hourly)
```
2. Which source has the greatest mean output by day? Which has the least? (Do not include zero values.)


```{r}
long_ca_energy %>%
  filter(output>0) %>% 
  group_by(day, source) %>%
  summarize(mean_daily = mean(output, na.rm = T)) %>%
  arrange(desc(mean_daily))
```

3. Which sources has the greatest variance in usage over the course of a dataset? Which has the least? (Do not include zero values.)

```{r}
long_ca_energy %>%
  filter(output>0) %>%
  group_by(source) %>%
  summarize(sd_output = sd(output, na.rm = T)) %>%
  arrange(desc(sd_output))
```



## Analyzing renewable versus non-renewable energy sources

The dataset `regroup.csv` has information about which sources are considered renewable by the state of California. Use this dataset, along with yourdata manipulation skills, to explore the use of renewable and non-renewable sources. Annotate what your descisions for the analysis.

Hint: Use your merge skills to merge the CA energy data with the `regroup` data. Which variable should you join by?

```{r}
regroup <- read.csv(here::here("data/ca_energy_regroup.csv"), 
                          stringsAsFactors = F)

head(regroup)



```

